\documentclass[10pt, conference, compsocconf]{IEEEtran}
% Set letter paper size:
\setlength{\paperheight}{11in}
\setlength{\paperwidth}{8.5in}
\usepackage[pass]{geometry}
\usepackage{amsmath}

\usepackage{color}
\usepackage{tikz}
\usetikzlibrary{arrows,positioning} 

\usepackage{minted}
\usepackage[textsize=footnotesize]{todonotes}

\begin{document}
%
% --- Author Metadata here ---
%\conferenceinfo{SC'15}{'15 Austin, Texas USA}
%\CopyrightYear{2007} % Allows default copyright year (20XX) to be over-ridden - IF NEED BE.
%\crdata{0-12345-67-8/90/01}  % Allows default copyright data (0-89791-88-6/97/05) to be over-ridden - IF NEED BE.
% --- End of Author Metadata ---

\title{Increasing the Energy Efficiency of Distributed Sorting by Avoiding Communication}

\author{\IEEEauthorblockN{Hari Sundar, 
Saurav Muralidharan, Amit Roy, Mary Hall, Prajakta Mane}
\IEEEauthorblockA{University of Utah\\
Salt Lake City, UT\\
E-mail: \{hari, sauravm, aroy, mhall, mane\}@cs.utah.edu}
}

\maketitle
\begin{abstract}
Sorting is an essential building block for several algorithms and applications. As we ramp up to the first exascale systems, it is important to look beyond the usual metrics of performance and scalability and also consider the energy and power requirements of sorting algorithms. 
In this work we improve the energy efficiency of distributed sorting by proposing a new algorithm that avoids communication whenever possible. In addition, we analyze the most energy and power efficient sorting strategies on GPUs and Heterogeneous system architectures. 
\end{abstract}

% \category{H.4}{Information Systems Applications}{Miscellaneous}
% \category{D.2.8}{Software Engineering}{Metrics}[complexity measures, performance measures]

% \terms{Theory}

% \keywords{ACM proceedings, \LaTeX, text tagging}

\section{Introduction}

Large-scale distributed sorting have focused on improving sort throughput (number of keys sorted per sec) and parallel scalability. However, for the next generation of supercomputers, additional metrics related to energy and power consumption are going to be equally if not more important. Since, sorting is entirely dominated by data movement, energy is equivalent to the total data communicated and power is proportional to the peak bandwidth utilized. This makes distributed sorting an ideal algorithm to study energy and power related optimizations, at least for memory-dominated applications. The motivation is to develop a distributed sorting algorithm that aims to minimize the overall data-movement and the peak bandwidth required with minimal sacrifices on performance or scalability. The development of such a large-scale distributed sorting algorithm is the central contribution of this work. We propose both distributed as well as node-local strategies to achieve this goal. Our expectation is that this work will serve as guideline for the development of such energy and power efficient for a wider array of problems in the near future.

\paragraph{Relevance} Why is this work relevant. 
 
Motivate and connect to a larger set of problems. Highlight contributions and list limitations.

\input{background.tex}
\input{distsort.tex}
\input{nodesort.tex}
\input{experiments.tex}
\input{results.tex}

\section{Discussion}
Discuss results and what the key findings are and what it means for future architectures.


%ACKNOWLEDGMENTS are optional
%\section{Acknowledgments}
%Thank Christopher Strong, nvidia?, Funding agencies.

\input{conclusions}

%
\bibliographystyle{IEEEtran}
\bibliography{esort,nitro,tuning} 

%\balancecolumns % GM June 2007
% That's all folks!
\end{document}
